Higher-Order unification for free!

### 2024-04-12 paper/main.tex
reword

  The object logic we are interested in is % an extension of the latter,
  Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
  we want to implement a unification based 
  proof search procedure using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi also comes with an encoding for CIC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only.

  In this paper we propose a better-behaved encoding, \Ho,
  and show how to map unification problems in \Fo{}
  to related problems in \Ho.
    As a result we can lift the ML
  higher-order unification procedure to \Fo{} in a way that it
  honours $\eta\beta$-equivalence (for CIC functions), solves
  problems in the pattern fragment and finally allows
  for the addition of heuristics to deal with problems outside the
  pattern fragment. The unification procedure we build can be used to
  effectively simulate a logic program in \Fo{} by re-using optimizations
  provided by the ML such as clause indexing.

@@

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a designated database of ``type class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper we focus on one aspect of
this work, precisely \emph{how to reuse the higher order unification procedure
of the meta language in order to simulate a higher order logic program
for the object language}.

@@

>>>>

  The object logic we are interested in is % an extension of the latter,
  Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
  we want to implement a higher-order unification-based 
  proof search procedure using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher-order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi comes with an encoding of CIC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but restricts \Ue to roughly first-order unification problems only.
  We call this basic encoding \Fo.

  In this paper we propose a better-behaved encoding \Ho,
  and show how to map unification problems in \Fo{}
  to related problems in \Ho.
    As a result we can lift the ML
  higher-order unification procedure to \Fo{} in a way that it
  honours $\eta\beta$-equivalence (for CIC functions), solves
  problems in the pattern fragment and finally allows
  for the addition of heuristics to deal with problems outside the
  pattern fragment. The unification procedure we build can be used to
  efficiently simulate a logic program in \Fo{} by re-using optimizations
  provided by the ML such as clause indexing.

@@

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a designated database of ``type class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

@@

In this paper we show how to lift the meta language higher-order
unification procedure to the object language. The resulting
unification honours $\eta\beta$-equivalence for
